---
title: "mos_w3"
author: "Jingsong Zhou"
date: "2023-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# read the dataset
dataset <- read.table("my_agam_aging_expr_timeseries_Jun23_2023.tsv", header = TRUE)
```


### 1. Variance Analysis of Single Gene Expression over Time

In this analysis, we investigate whether there is evidence of variance in the expression of a single gene increasing with time. We iterate over a selected set of geneIDs and compute the variance of gene expression for each age group. This analysis helps us understand if the regulatory robustness of gene expression changes with age, which has implications for evolutionary perspectives on gene regulation.

```{r}
# Get unique geneIDs from the dataset
geneIDs <- unique(dataset$geneID)

# Split the dataset by geneID
gene_data_list <- split(dataset, dataset$geneID)
```

The goal of this analysis is to determine whether there is evidence of variance in the expression of a single gene increasing or decreasing with time. By examining the variance of gene expression at different ages and fitting a linear regression model, we can assess the significance of the age coefficient and determine if there is a positive/negative trend in variance over time.

```{r}
# Initialize the variables to count genes with changing trend
genes_with_increasing_variance <- 0
genes_with_decreasing_variance <- 0
increasing_trend_age_genes<-c()
decreasing_trend_age_genes<-c()

# Initialize an empty vector to store the coefficients
coefficients_vector <- c()

# Iterate over the gene IDs
for (i in 1:length(geneIDs)) {
  gene <- geneIDs[i]
  
  # Subset the dataset for the specific geneID
  gene_data <- gene_data_list[[as.character(gene)]]
  
  # Compute the variance of gene expression for each age
  variance_by_age <- aggregate(transcription ~ age, data = gene_data, FUN = var)
  
  # Fit a linear regression model
  lm_model <- lm(variance_by_age$transcription ~ variance_by_age$age)
  
  # Check the significance of the age coefficient
  p_value <- summary(lm_model)$coefficients[2, 4]

  # Check the sign of the age coefficient
  coefficient <- summary(lm_model)$coefficients[2, 1]
  
  # Store the coefficient in the vector
  coefficients_vector <- c(coefficients_vector, coefficient)  
  
  # Make a conclusion based on the p-value and coefficient sign
  if (p_value < 0.05 & coefficient > 0) {
    # There is significant evidence of increasing variance over ages
    genes_with_increasing_variance <- genes_with_increasing_variance + 1
    increasing_trend_age_genes<-c(increasing_trend_age_genes, gene)
  }
  else if (p_value < 0.05 & coefficient < 0) {
    # There is significant evidence of decreasing variance over ages
    genes_with_decreasing_variance <- genes_with_decreasing_variance + 1
    decreasing_trend_age_genes<-c(decreasing_trend_age_genes, gene)
  }
}
# Calculate the proportion of genes with increasing variance
proportion_increasing_variance <- genes_with_increasing_variance / length(geneIDs)
# Calculate the proportion of genes with decreasing variance
proportion_decreasing_variance <- genes_with_decreasing_variance / length(geneIDs)

# Print the summary result
cat("Proportion of genes with increasing variance over time:", proportion_increasing_variance, "\n")
cat("Proportion of genes with decreasing variance over time:", proportion_decreasing_variance, "\n")
```

The code below generates a histogram to visualize the distribution of coefficients. The coefficients represent the relationship between age and gene expression variance for each gene. By examining the distribution, we can gain insights into the overall pattern and variability of the coefficients. This information helps us understand the extent to which gene expression variance changes with age.

```{r}
# Plot the distribution of coefficients
hist(coefficients_vector, main = "Distribution of Coefficients",
     xlab = "Coefficient", ylab = "Proportion", freq = FALSE)
```

The code creates a quantile-quantile (QQ) plot to assess the normality of the coefficients. The QQ plot compares the observed distribution of coefficients against the expected distribution under normality. If the points on the plot roughly align with the diagonal line (qqline), it suggests that the coefficients follow a normal distribution. Departures from the diagonal line indicate deviations from normality, which can be important for statistical analyses that assume normality.

```{r}
# Create a quantile-quantile (QQ) plot
qqnorm(coefficients_vector)
qqline(coefficients_vector)
```

This section performs bootstrap resampling to estimate the uncertainty associated with a symmetry measure of the coefficients. The symmetry measure, in this case, is the mean of the bootstrap samples. We use Bootstrap method to resample the coefficients with replacement, calculate the mean of each bootstrap sample, and store it in the bootstrap_samples vector. The resulting vector represents a distribution of bootstrap symmetry measures. The confidence interval for the symmetry measure is then computed using the quantile function, specifying the desired percentiles. Finally, the confidence interval is printed to provide an estimate of the uncertainty surrounding the symmetry measure of the coefficients.

```{r}
# Set the number of bootstrap iterations
num_iterations <- 1000

# Initialize an empty vector to store bootstrap samples
bootstrap_samples <- numeric(num_iterations)

# Perform bootstrapping
for (i in 1:num_iterations) {
  # Resample the coefficients with replacement
  bootstrap_sample <- sample(coefficients_vector, replace = TRUE)
  
  # Calculate the symmetry measure (mean) of the bootstrap sample
  bootstrap_symmetry <- mean(bootstrap_sample)  # Replace with the symmetry measure
  
  # Store the symmetry measure in the bootstrap_samples vector
  bootstrap_samples[i] <- bootstrap_symmetry
}

# Calculate the confidence interval for the symmetry measure
confidence_interval <- quantile(bootstrap_samples, c(0.025, 0.975))

# Print the confidence interval
cat("Bootstrap Confidence Interval:", confidence_interval, "\n")
```

For a more intuitive understanding, we use visualization methods for genes with increasing or decreasing trend in gene expression variance to observe the change in variance with age.

```{r}
# Set the maximum number of iterations
max_iterations <- 9  

# Set the number of rows and columns for the grid
num_rows <- 3
num_cols <- 3
# Create a new blank plot
plot.new()
# Set up the grid layout
par(mfrow = c(num_rows, num_cols), mar = c(4, 4, 2, 2))

for (i in 1:max_iterations) {
  gene <- increasing_trend_age_genes[i]
  
  # Subset the dataset for the specific geneID
  gene_data <- gene_data_list[[as.character(gene)]]
  
  # Compute the variance of gene expression for each age
  variance_by_age <- aggregate(transcription ~ age, data = gene_data, FUN = var)
  
  # Plot the variance by age
  plot(variance_by_age$age, variance_by_age$transcription, type = "b", 
       xlab = "Age", ylab = "Expression Variance",
       main = paste0("Gene ID:", gene))
}

# Reset the plot layout
par(mfrow = c(1, 1))
```


```{r}
# Set the maximum number of iterations
max_iterations <- 9  

# Set the number of rows and columns for the grid
num_rows <- 3
num_cols <- 3
# Create a new blank plot
plot.new()
# Set up the grid layout
par(mfrow = c(num_rows, num_cols), mar = c(4, 4, 2, 2))

for (i in 1:max_iterations) {
  gene <- decreasing_trend_age_genes[i]
  
  # Subset the dataset for the specific geneID
  gene_data <- gene_data_list[[as.character(gene)]]
  
  # Compute the variance of gene expression for each age
  variance_by_age <- aggregate(transcription ~ age, data = gene_data, FUN = var)
  
  # Plot the variance by age
  plot(variance_by_age$age, variance_by_age$transcription, type = "b", 
       xlab = "Age", ylab = "Gene Expression Variance",
       main = paste0("Gene ID:", gene))
}

# Reset the plot layout
par(mfrow = c(1, 1))
```

### 2. Correlation Analysis of Gene Expression between Different Genes over Time

The code analyzes the correlation between two selected genes over different ages. It subsets the dataset to extract the gene data, calculates the correlation coefficient for each age, and plots the correlation coefficients over time. This analysis helps to understand how the correlation between the genes changes with age, providing insights into gene interactions and potential biological processes related to aging or development.

```{r}
correlation_genes_over_ages <- function(selected_gene1,selected_gene2){ #This function computes the correlation for the chosen genes pair over ages and return a vector containing the correlation
  
  # Subset the dataset for the selected genes
  selected_gene1_data <- gene_data_list[[as.character(selected_gene1)]]
  selected_gene2_data <- gene_data_list[[as.character(selected_gene2)]]
  
  # Initialize vectors to store correlation coefficients and ages
  correlation_coefficients <- c()
  
  # Iterate over unique ages
  for (i in unique(dataset$age)) {
    # Subset the data for the current age
    subset_gene1_data <- subset(selected_gene1_data, age == i)
    subset_gene2_data <- subset(selected_gene2_data, age == i)
    
    # Check if there are multiple measurements for the same experimentID
    if (length(unique(dataset$experimentID)) != nrow(selected_gene1_data)) {
      # Calculate the average transcription value for each experimentID
      subset_gene1_data <- aggregate(transcription ~ experimentID, data = subset_gene1_data, FUN = mean)
    }
    if (length(unique(dataset$experimentID)) != nrow(selected_gene2_data)) {
      # Calculate the average transcription value for each experimentID
      subset_gene2_data <- aggregate(transcription ~ experimentID, data = subset_gene2_data, FUN = mean)
    }
    
    # Merge the data based on experimentID
    merged_data <- merge(subset_gene1_data, subset_gene2_data, by = "experimentID")
    
    # Extract the gene1 and gene2 data for the current age
    gene1_data <- merged_data$transcription.x
    gene2_data <- merged_data$transcription.y

    # Calculate the correlation coefficient
    correlation <- cor(gene1_data, gene2_data)
    
    # Store the correlation coefficient and age
    correlation_coefficients <- c(correlation_coefficients, correlation)
  }
  return(correlation_coefficients)
}
```

The code performs a correlation analysis between randomly selected pairs of genes over different ages. The resulting plots provide insights into the variability and patterns of gene-gene correlations across different gene pairs.

```{r}
# Set the maximum number of iterations
max_iterations <- 9  

# Set the number of rows and columns for the grid
num_rows <- 3
num_cols <- 3
# Create a new blank plot
plot.new()
# Set up the grid layout
par(mfrow = c(num_rows, num_cols), mar = c(4, 4, 2, 2))

for (i in 1:max_iterations) {
  # Randomly select a pair of genes for correlation analysis
  random_genes <- sample(geneIDs, size = 2)  # we can adjust the size as desired
  # Compute the correlation
  correlation_coefficients <- correlation_genes_over_ages(random_genes[1],random_genes[2])
  # Plot the correlation coefficients over ages
  plot(unique(dataset$age), correlation_coefficients,type = 'b', pch = 19, xlab = "Age", ylab = "Correlation Coefficient", main = "Correlation Coefficient over Ages")
}

# Reset the plot layout
par(mfrow = c(1, 1))
```

The goal of this analysis is to determine whether there is evidence of correlation between the expression of two genes increasing or decreasing with time. By examining the correlation of gene expression at different ages and fitting a linear regression model, we can assess the significance of the age coefficient and determine if there is a positive/negative trend in correlation over time.

In the below code, we randomly select gene pairs and calculate the correlation between their expression over age. We then fit a linear regression model to determine the trend based on the slope of the regression line. Gene pairs with positive slopes indicate increasing trends, while gene pairs with negative slopes indicate decreasing trends. The gene pairs with increasing and decreasing trends are stored in separate vectors.

```{r}
# Initialize the variables to count genes with changing trend
genes_with_increasing_correlation <- 0
genes_with_decreasing_correlation <- 0
ages <- unique(dataset$age)

# Initialize an empty vector to store the coefficients
cor_coefficients_vector <- c()

# Set the maximum number of iterations
max_iterations <- 200

for (i in 1:max_iterations) {
  
    # Randomly select a pair of genes for correlation analysis
    random_genes <- sample(geneIDs, size = 2)
    
    # Compute the correlation of gene expression for each age
    correlation_coefficients <- correlation_genes_over_ages(random_genes[1],random_genes[2])  
  
    # Fit a linear regression model
    lm_model <- lm(correlation_coefficients ~ ages)

    # Check the significance of the age coefficient
    p_value <- summary(lm_model)$coefficients[2, 4]

    # Check the sign of the age coefficient
    coefficient <- summary(lm_model)$coefficients[2, 1]
  
    # Store the coefficient in the vector
    cor_coefficients_vector <- c(cor_coefficients_vector, coefficient)  
  
    # Make a conclusion based on the p-value and coefficient sign
    if (p_value < 0.1 & coefficient > 0) {
      # There is significant evidence of increasing correlation over ages
     genes_with_increasing_correlation <- genes_with_increasing_correlation + 1
    }
    else if (p_value < 0.1 & coefficient < 0) {
     # There is significant evidence of decreasing correlation over ages
     genes_with_decreasing_correlation <- genes_with_decreasing_correlation + 1
    }
}

# Calculate the proportion of genes with increasing correlation
proportion_increasing_correlation <- genes_with_increasing_correlation / max_iterations

# Calculate the proportion of genes with decreasing variance
proportion_decreasing_correlation <- genes_with_decreasing_correlation / max_iterations

# Print the summary result
cat("Proportion of genes with increasing correlation over time:", proportion_increasing_correlation, "\n")
cat("Proportion of genes with decreasing correlation over time:", proportion_decreasing_correlation, "\n")

# Plot the distribution of coefficients
hist(coefficients_vector, main = "Distribution of Coefficients",
     xlab = "Coefficient", ylab = "Proportion", freq = FALSE)
```




```{r}
# Create a quantile-quantile (QQ) plot
qqnorm(coefficients_vector)
qqline(coefficients_vector)
```









