---
title: "mos_w2"
author: "Jingsong Zhou"
date: "2023-06-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# read the dataset
dataset <- read.table("my_agam_aging_expr_timeseries_Jun23_2023.tsv", header = TRUE)
```


1. Variance Analysis of Single Gene Expression over Time

In this analysis, we investigate whether there is evidence of variance in the expression of a single gene increasing with time. We iterate over a selected set of geneIDs and compute the variance of gene expression for each age group. This analysis helps us understand if the regulatory robustness of gene expression changes with age, which has implications for evolutionary perspectives on gene regulation.

```{r}
# Get unique geneIDs from the dataset
geneIDs <- unique(dataset$geneID)

# Split the dataset by geneID
gene_data_list <- split(dataset, dataset$geneID)
```

The goal of this analysis is to determine whether there is evidence of variance in the expression of a single gene increasing or decreasing with time. By examining the variance of gene expression at different ages and fitting a linear regression model, we can assess the significance of the age coefficient and determine if there is a positive/negative trend in variance over time.

```{r}
# Initialize the variables to count genes with changing trend
genes_with_increasing_variance <- 0
genes_with_decreasing_variance <- 0
var_trend_age_genes<-c()

# Initialize an empty vector to store the coefficients
coefficients_vector <- c()

# Iterate over the gene IDs
for (i in 1:length(geneIDs)) {
  gene <- geneIDs[i]
  
  # Subset the dataset for the specific geneID
  gene_data <- gene_data_list[[as.character(gene)]]
  
  # Compute the variance of gene expression for each age
  variance_by_age <- aggregate(transcription ~ age, data = gene_data, FUN = var)
  
  # Fit a linear regression model
  lm_model <- lm(log(variance_by_age$transcription) ~ variance_by_age$age)
  
  # Check the significance of the age coefficient
  p_value <- summary(lm_model)$coefficients[2, 4]

  # Check the sign of the age coefficient
  coefficient <- summary(lm_model)$coefficients[2, 1]
  
  # Store the coefficient in the vector
  coefficients_vector <- c(coefficients_vector, coefficient)  
  
  # Make a conclusion based on the p-value and coefficient sign
  if (p_value < 0.05 & coefficient > 0) {
    # There is significant evidence of increasing variance over ages
    genes_with_increasing_variance <- genes_with_increasing_variance + 1
    var_trend_age_genes<-c(var_trend_age_genes, gene)
  }
  else if (p_value < 0.05 & coefficient < 0) {
    # There is significant evidence of decreasing variance over ages
    genes_with_decreasing_variance <- genes_with_decreasing_variance + 1
  }
}
# Calculate the proportion of genes with increasing variance
proportion_increasing_variance <- genes_with_increasing_variance / length(geneIDs)
# Calculate the proportion of genes with decreasing variance
proportion_decreasing_variance <- genes_with_decreasing_variance / length(geneIDs)

# Print the summary result
cat("Proportion of genes with increasing variance over time:", proportion_increasing_variance, "\n")
cat("Proportion of genes with decreasing variance over time:", proportion_decreasing_variance, "\n")
```

```{r}
# Plot the distribution of coefficients
hist(coefficients_vector, main = "Distribution of Coefficients",
     xlab = "Coefficient", ylab = "Frequency")
```

```{r}
# Create a quantile-quantile (QQ) plot
qqnorm(coefficients_vector)
qqline(coefficients_vector)
```

```{r}
# Set the number of bootstrap iterations
num_iterations <- 1000

# Initialize an empty vector to store bootstrap samples
bootstrap_samples <- numeric(num_iterations)

# Perform bootstrapping
for (i in 1:num_iterations) {
  # Resample the coefficients with replacement
  bootstrap_sample <- sample(coefficients_vector, replace = TRUE)
  
  # Calculate the symmetry measure (e.g., mean, median) of the bootstrap sample
  bootstrap_symmetry <- mean(bootstrap_sample)  # Replace with your symmetry measure
  
  # Store the symmetry measure in the bootstrap_samples vector
  bootstrap_samples[i] <- bootstrap_symmetry
}

# Calculate the confidence interval for the symmetry measure
confidence_interval <- quantile(bootstrap_samples, c(0.025, 0.975))

# Print the confidence interval
cat("Bootstrap Confidence Interval:", confidence_interval, "\n")
```





For a more intuitive understanding, we use visualization methods for genes with an increasing trend in gene expression variance to observe the change in variance with age.

```{r}
# Set the maximum number of iterations
max_iterations <- 20  

# Set the number of rows and columns for the grid
num_rows <- 4
num_cols <- 5
# Create a new blank plot
plot.new()
# Set up the grid layout
par(mfrow = c(num_rows, num_cols), mar = c(2, 2, 2, 2))

for (i in 1:max_iterations) {
  gene <- var_trend_age_genes[i]
  
  # Subset the dataset for the specific geneID
  gene_data <- gene_data_list[[as.character(gene)]]
  
  # Compute the variance of gene expression for each age
  variance_by_age <- aggregate(transcription ~ age, data = gene_data, FUN = var)
  
  # Plot the variance by age
  plot(variance_by_age$age, log(variance_by_age$transcription), type = "b", 
       xlab = "Age", ylab = "Gene Expression Variance",
       main = paste0("Gene ID:", gene))
}

# Reset the plot layout
par(mfrow = c(1, 1))
```

2. Correlation Analysis of Gene Expression between Different Genes over Time

In this analysis, we examine the correlation of expression between different genes over time. We select a set of geneIDs of interest and calculate the correlation coefficients based on their expression levels at each age point. The correlation coefficients are then visualized, where each point represents the correlation between a pair of genes. This analysis allows us to explore the relationships and potential co-regulation patterns between different genes.

Here we use three methods to select geneIDs for correlation analysis.

(a)Random Sampling:

```{r}
library(corrplot)
# Randomly select a subset of genes
random_genes <- sample(geneIDs, size = 10)  # we can adjust the size as desired

# Subset the dataset for the selected geneIDs
gene_data_subset <- subset(dataset, geneID %in% random_genes)

# Calculate the average or median transcription values for each geneID at each time point
averages_by_time <- aggregate(transcription ~ geneID + age, data = gene_data_subset, FUN = mean)

# Reshape the data frame to have time points as columns and geneIDs as rows
correlation_data <- reshape(averages_by_time, idvar = "age", timevar = "geneID", direction = "wide")
# Remove the geneID column
correlation_data <- correlation_data[, -1]  

colnames(correlation_data)<-gsub('transcription.','',colnames(correlation_data))

# Calculate the correlation matrix for the selected geneIDs
correlation_matrix <- cor(correlation_data)

# Create a correlogram
corrplot(correlation_matrix, method = "circle", type = "full",
         tl.col = "black", tl.srt = 30, tl.cex = 0.5)

```


(b)Top-Ranked Genes:

```{r}
# Calculate variance for each gene
gene_variance <- aggregate(transcription ~ geneID, data = dataset, FUN = var)

# Sort genes by variance in descending order
top_genes <- head(gene_variance[order(gene_variance$transcription, decreasing = TRUE), "geneID"], n = 10)  # Adjust the number of top genes as desired

# Subset the dataset for the selected geneIDs
gene_data_subset <- subset(dataset, geneID %in% top_genes)

# Calculate the average or median transcription values for each geneID at each time point
averages_by_time <- aggregate(transcription ~ geneID + age, data = gene_data_subset, FUN = mean)

# Reshape the data frame to have time points as columns and geneIDs as rows
correlation_data <- reshape(averages_by_time, idvar = "age", timevar = "geneID", direction = "wide")
# Remove the geneID column
correlation_data <- correlation_data[, -1]  

colnames(correlation_data)<-gsub('transcription.','',colnames(correlation_data))

# Calculate the correlation matrix for the selected geneIDs
correlation_matrix <- cor(correlation_data)

# Create a correlogram
corrplot(correlation_matrix, method = "circle", type = "full",
         tl.col = "black", tl.srt = 30, tl.cex = 0.5)
```


(c)Genes with increasing expression variance over ages:

```{r}
# Randomly select a subset of genes
increasing_genes <- sample(var_trend_age_genes, size = 10)  

# Subset the dataset for the selected geneIDs
gene_data_subset <- subset(dataset, geneID %in% increasing_genes)

# Calculate the average or median transcription values for each geneID at each time point
averages_by_time <- aggregate(transcription ~ geneID + age, data = gene_data_subset, FUN = mean)

# Reshape the data frame to have time points as columns and geneIDs as rows
correlation_data <- reshape(averages_by_time, idvar = "age", timevar = "geneID", direction = "wide")
# Remove the geneID column
correlation_data <- correlation_data[, -1]  

colnames(correlation_data)<-gsub('transcription.','',colnames(correlation_data))

# Calculate the correlation matrix for the selected geneIDs
correlation_matrix <- cor(correlation_data)

# Create a correlogram
corrplot(correlation_matrix, method = "circle", type = "full",
         tl.col = "black", tl.srt = 30, tl.cex = 0.5)

```






